---
title: "Astrocyte RNA-seq – ATP13A2 (KOLF c.1306 vs WT)"
author: "E. Coccia"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: true
params:
  data_dir: "data"
  out_dir: "outputs"
  project_seed: 123
---

## Overview

This document performs differential gene expression analysis of astrocyte RNA-seq data comparing ATP13A2 c.1306 KOLF lines versus WT controls.

### Main analysis steps
1. Load transcript-level Kallisto counts
2. Parse transcript annotation fields
3. Aggregate transcript counts to gene-level counts
4. Construct sample group labels from sample names
5. Perform filtering and TMM normalization with edgeR
6. Generate PCA from logCPM values
7. Perform quasi-likelihood differential expression testing with edgeR
8. Generate a volcano plot highlighting lysosomal and neuroinflammation-related genes
9. Perform GO Molecular Function enrichment analysis with clusterProfiler

### Figures generated from this script
- **Figure 2B**: Volcano plot highlighting lysosomal and neuroinflammation genes
- **Figure S2A**: GO Molecular Function dot plot (clusterProfiler)

### Software and packages
This analysis uses R packages including:
- tidyverse
- edgeR
- ggrepel
- fs
- readr
- ggplot2
- here
- clusterProfiler
- org.Hs.eg.db
- patchwork

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 5
)
set.seed(params$project_seed)

suppressPackageStartupMessages({
  library(tidyverse)
  library(edgeR)
  library(ggrepel)
  library(fs)
  library(readr)
  library(ggplot2)
  library(here)
})
```
```{r setup, include=FALSE}
# ---- load transcript-level counts ----
load(here::here(params$data_dir, "resultsgencodev38_kallisto_counts.RData"))
stopifnot(exists("tx_counts"))

tx_counts <- as.data.frame(tx_counts) %>%
  tibble::rownames_to_column("transcript")

# Expected transcript rowname format:
# ENST|ENSG|havana_gene|havana_tx|transcript_name|gene_name|x|transcript_type

parsed <- tx_counts %>%
  tidyr::separate(
    col  = transcript,
    into = c("transcript_id","gene_id","havana_gene","havana_transcript",
             "transcript_name","gene_name","x","transcript_type"),
    sep  = "\\|", fill = "right", remove = FALSE
  ) %>%
  tidyr::drop_na(gene_name)

# ---- aggregate to gene counts ----
gene_counts <- parsed %>%
  dplyr::select(gene_name, where(is.numeric)) %>%
  dplyr::group_by(gene_name) %>%
  dplyr::summarise(dplyr::across(everything(), sum, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(gene_name)

# ---- build group labels from column names ----
groups <- dplyr::case_when(
  stringr::str_detect(keep_cols, regex("c\\.1306", ignore_case = TRUE)) ~ "ATP13A2_c.1306",
  stringr::str_detect(keep_cols, regex("\\bWT\\b|_WT", ignore_case = TRUE)) ~ "WT",
  TRUE ~ "UNKNOWN"
)
```

```{r setup, include=FALSE}
# ---- DGEList, filtering, normalization ----
kolf_counts <- gene_counts_kolf %>%
  tibble::column_to_rownames("gene_name") %>%
  as.matrix()

d_KOLF <- edgeR::DGEList(
  counts = kolf_counts,
  group  = factor(groups, levels = c("WT","ATP13A2_c.1306"))
)

keep <- rowSums(edgeR::cpm(d_KOLF) > 10) >= 2
d_KOLF <- d_KOLF[keep, , keep.lib.sizes = FALSE]

d_KOLF$samples$lib.size <- colSums(d_KOLF$counts)
d_KOLF <- edgeR::calcNormFactors(d_KOLF, method = "TMM")

logCPM <- edgeR::cpm(d_KOLF, log = TRUE, prior.count = 1)

# ---- PCA ----
pca <- prcomp(t(logCPM), scale. = FALSE)
var_expl <- (pca$sdev^2 / sum(pca$sdev^2)) * 100

pca_df <- tibble::as_tibble(pca$x, rownames = "sample") %>%
  dplyr::mutate(condition = d_KOLF$samples$group[match(sample, rownames(d_KOLF$samples))])

p_pca <- ggplot2::ggplot(pca_df, ggplot2::aes(PC1, PC2, color = condition, label = sample)) +
  ggplot2::geom_point(size = 3) +
  ggrepel::geom_text_repel(show.legend = FALSE) +
  ggplot2::labs(
    title = "PCA (edgeR logCPM) – KOLF c.1306 vs WT",
    x = sprintf("PC1 (%.1f%%)", var_expl[1]),
    y = sprintf("PC2 (%.1f%%)", var_expl[2]),
    color = "Condition"
  ) +
  ggplot2::theme_bw()

p_pca
save_fig(p_pca, "pca_kolf_edger", width = 7.5, height = 5.5)
```

```{r setup, include=FALSE}

# Use the filtered and normalized DGEList from PCA step
stopifnot(exists("d_KOLF"))
d_KOLF1 <- d_KOLF

designKOLF <- model.matrix(~ 0 + group, data = d_KOLF1$samples)
colnames(designKOLF) <- levels(d_KOLF1$samples$group)
designKOLF

d_KOLF1 <- estimateDisp(d_KOLF1, designKOLF)

png(here::here(params$out_dir, "figures", "BCV_KOLF.png"), width = 1600, height = 1200, res = 150)
plotBCV(d_KOLF1, main = "BCV – KOLF (edgeR QL)")
dev.off()

fitK <- glmQLFit(d_KOLF1, designKOLF)
contrastK <- makeContrasts(ATP13A2_c.1306 - WT, levels = designKOLF)
qlfK <- glmQLFTest(fitK, contrast = contrastK)

RNAseq_DEGs <- topTags(qlfK, n = Inf)$table %>%
  tibble::rownames_to_column("gene") %>%
  arrange(FDR)

readr::write_csv(RNAseq_DEGs, ("RNAseq_DEGs.csv"))

# DEG numbers report summary
fc_cutoff <- 1
fdr_cutoff <- 0.01

sig_genes <- RNAseq_DEGs %>%
  filter(!is.na(FDR), FDR < fdr_cutoff, abs(logFC) >= fc_cutoff)

n_up <- sig_genes %>% filter(logFC >= fc_cutoff) %>% nrow()
n_down <- sig_genes %>% filter(logFC <= -fc_cutoff) %>% nrow()

n_up
n_down

# ------ FIGURE 2B----------
RNAseq_DEG <- df
library(dplyr)
library(ggplot2)
library(ggrepel)


# Gene vectors
lyso_genes  <- LYSOSOME$gene
neuro_genes <- Neuroinflammation$`Molecule Name`

# Fold-change cutoff
fc_cutoff <- 0.7

# Prepare DE table
df <- RNAseq_DEGs %>%
  filter(!is.na(FDR), FDR > 0, !is.na(logFC)) %>%
  mutate(
    is_lyso  = gene %in% lyso_genes,
    is_neuro = gene %in% neuro_genes,
    highlight = (is_lyso | is_neuro) & abs(logFC) >= fc_cutoff,
    label_color = case_when(
      is_lyso  ~ "aquamarine4",
      is_neuro ~ "darkorange3",
      TRUE     ~ NA_character_
    )
  )

ggplot(df, aes(x = logFC, y = -log10(FDR))) +
  # Cutoff lines
  geom_hline(yintercept = -log10(0.01),
             linetype = "dotted", color = "azure3", linewidth = 0.6) +
  geom_vline(xintercept = c(-fc_cutoff, fc_cutoff),
             linetype = "dotted", color = "azure3", linewidth = 0.6) +
  
  # Background genes
  geom_point(
    data = subset(df, !is_lyso & !is_neuro),
    color = "azure3",
    alpha = 0.35,
    size  = 1.8
  ) +
  
  # Neuroinflammation genes
  geom_point(
    data = subset(df, is_neuro),
    color = "darkorange1",
    alpha = 0.65,
    size  = 2.4
  ) +
  
  # Lysosome genes
  geom_point(
    data = subset(df, is_lyso),
    color = "aquamarine3",
    alpha = 0.65,
    size  = 2.4
  ) +
  
  # Labels + connectors (bigger text)
  geom_text_repel(
    data = subset(df, highlight),
    aes(label = gene),
    size = 4,                       
    color = df$label_color[df$highlight],
    box.padding = 0.5,
    point.padding = 0.4,
    segment.color = df$label_color[df$highlight],
    segment.size = 0.45,
    max.overlaps = Inf
  ) +
  
  labs(
    title = "Volcano plot: Neuroinflammation & Lysosome genes",
    x = "log2 Fold Change",
    y = expression(-log[10]("FDR"))
  ) +
  
  theme_bw(base_size = 16) +           
  theme(
    plot.title   = element_text(size = 22, face = "bold"),
    axis.title   = element_text(size = 20),
    axis.text    = element_text(size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.6)
  )


# ------FIGURE S2A----------
## ---- GO-MF-up-vs-down, message=FALSE, warning=FALSE --------------------
install.packages("tidytext")

library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
library(dplyr)

ego_up_mf <- enrichGO(gene          = up_genes,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = "SYMBOL",
                      ont           = "MF",
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.01,
                      readable      = TRUE)

ego_up_mf_simple <- simplify(ego_up_mf, 
                             cutoff = 0.7, 
                             by = "p.adjust", 
                             select_fun = min)

ego_down_mf <- enrichGO(gene          = down_genes,
                        OrgDb         = org.Hs.eg.db,
                        keyType       = "SYMBOL",
                        ont           = "MF",
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        readable      = TRUE)

ego_down_mf_simple <- simplify(ego_down_mf, 
                               cutoff = 0.7, 
                               by = "p.adjust", 
                               select_fun = min)

plot_up <- dotplot(ego_up_mf_simple, showCategory = 10) + 
  ggtitle("Simplified MF: Upregulated Genes")

plot_down <- dotplot(ego_down_mf_simple, showCategory = 10) + 
  ggtitle("Simplified MF: Downregulated Genes")

print(plot_up)
print(plot_down)

library(clusterProfiler)
library(ggplot2)
library(patchwork)

all_results <- rbind(as.data.frame(ego_up_mf_simple), 
                     as.data.frame(ego_down_mf_simple))

p_limits <- range(all_results$p.adjust)
count_limits <- range(all_results$Count)

p1 <- dotplot(ego_up_mf_simple, showCategory = 10) + 
  scale_color_gradient(low = "red", high = "blue", limits = p_limits) +
  scale_size_continuous(limits = count_limits) +
  ggtitle("Upregulated") +
  theme(legend.position = "none")

p2 <- dotplot(ego_down_mf_simple, showCategory = 10) + 
  scale_color_gradient(low = "red", high = "blue", limits = p_limits) +
  scale_size_continuous(limits = count_limits) +
  ggtitle("Downregulated")

combined_plot <- (p1 | p2) + 
  plot_layout(guides = "collect") +
  plot_annotation(title = "Molecular Function: Comparison of Top 10 Pathways",
                  subtitle = "Both plots share the same p-adjust and Count scales")

print(combined_plot)
```
