---
title: "Astrocyte ATAC-seq – ATP13A2 (WT vs c.1306)"
author: "E. Coccia"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: true
params:
  project_seed: 123
---

## Overview

This document analyzes differential chromatin accessibility in human iPSC-derived midbrain astrocytes comparing WT and ATP13A2 c.1306 mutant cells.

### Main analysis steps
1. Load annotated differential ATAC-seq peaks
2. Clean and format peak-level statistics
3. Simplify genomic annotations
4. Generate exploratory accessibility plots
5. Generate MA plots of differential accessibility
6. Focus on promoter-associated differential accessibility
7. Perform GO enrichment on differentially accessible promoter-associated genes
8. Integrate ATAC-seq with RNA-seq differential expression
9. Integrate ATAC-seq with DNA methylation data
10. Integrate RNA-seq with DNA methylation data

### Figures generated from this script
- **Figure 4I**: ATAC-seq / RNA-seq concordance scatterplot
- **Figure S4E**: ATAC-seq MA plot
- **Figure S4E**: Bar plots of differentially accessible promoters

### Software and packages
This analysis uses R packages including:
- tidyverse
- ggplot2
- ggrepel
- clusterProfiler
- org.Hs.eg.db
- enrichplot
- patchwork

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 5
)
set.seed(params$project_seed)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(ggrepel)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(enrichplot)
  library(patchwork)
})

# ---- load annotated differential ATACseq peaks ----
pseudo <- 1

atac_clean <- atac_raw %>%
  mutate(
    Fold               = as.numeric(as.character(Fold)),
    p.value            = as.numeric(as.character(p.value)),
    FDR                = as.numeric(as.character(FDR)),
    Conc_ATP13A2c1306  = as.numeric(as.character(Conc_ATP13A2c1306)),
    Conc_WT            = as.numeric(as.character(Conc_WT)),
    neglog10_p         = -log10(p.value),
    A = 0.5 * (log2(Conc_ATP13A2c1306 + pseudo) +
               log2(Conc_WT + pseudo)),
    M = log2((Conc_ATP13A2c1306 + pseudo) /
             (Conc_WT + pseudo))
  )

# ---- simplify genomic annotations ----
atac_clean <- atac_clean %>%
  mutate(
    annotation_simple = case_when(
      str_detect(annotation, regex("Promoter", ignore_case = TRUE)) ~ "Promoter",
      str_detect(annotation, regex("Intron", ignore_case = TRUE)) ~ "Intron",
      str_detect(annotation, regex("Exon|CDS|UTR|Gene Body", ignore_case = TRUE)) ~ "Gene body",
      str_detect(annotation, regex("Intergenic", ignore_case = TRUE)) ~ "Distal intergenic",
      TRUE ~ "Other"
    )
  )

ann_colors_simple <- c(
  "Promoter"          = "#F2D7A0",
  "Intron"            = "#5B6F96",
  "Gene body"         = "#A7CBE6",
  "Distal intergenic" = "#E89A9B",
  "Other"             = "gray70"
)

# ---- ild reduced peak table for exploratory analyses ----
atac_reduced <- atac_clean %>%
  filter(!is.na(gene_symbol)) %>%
  filter(!is.na(Fold), Fold > 0) %>%
  select(gene_symbol, Fold, p.value, FDR, annotation, annotation_simple) %>%
  mutate(log2FC = log2(Fold)) %>%
  relocate(log2FC, .after = Fold)
```

```{r setup, include=FALSE}
# ---- exploratory differential accessibile plots ----
## volcano plot genoimic annotations
ggplot(atac_reduced,
       aes(x = log2FC,
           y = -log10(p.value),
           color = annotation_simple)) +
  geom_point(alpha = 0.6, size = 1.3) +
  scale_color_manual(
    values = c(
      "Promoter"          = "#e41a1c",
      "Intron"            = "#377eb8",
      "Gene body"         = "#4daf4a",
      "Distal intergenic" = "#984ea3",
      "Other"             = "grey70"
    )
  ) +
  theme_minimal(base_size = 14) +
  labs(
    title = "ATAC-seq Differential Accessibility",
    x = "log2 Fold Change (ATP13A2 c.1306 vs WT)",
    y = "-log10 p-value",
    color = "Region"
  )

# ------ MA PLOT ----------
ggplot(atac_clean, aes(x = A, y = M)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  labs(
    title = "MA plot – ATAC-seq WT vs ATP13A2 c.1306",
    x = "A = mean accessibility (log2)",
    y = "M = log2 fold change"
  )


#--------FIGURE S4E: MA PLOT--------
ggplot(atac_clean, aes(x = A, y = M, color = annotation_simple)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = ann_colors_simple) +
  theme_bw() +
  labs(
    title = "MA plot – ATAC-seq WT vs ATP13A2 c.1306",
    x = "A = mean accessibility (log2)",
    y = "M = log2 fold change",
    color = "Annotation"
  )
#--------FIGURE S4E: ANNOTATION COMPOSITION OF PEAKS MORE AVAILABLE IN ATP13A2 C.1306--------
ann_pct_mut_simple <- atac_clean %>%
  filter(M > 0) %>%
  count(annotation_simple, name = "n") %>%
  mutate(pct = 100 * n / sum(n)) %>%
  mutate(annotation_simple = factor(annotation_simple, levels = names(ann_colors_simple)))

ann_pct_mut_simple

#--------FIGURE S4E: ANNOTATION COMPOSITION OF PEAKS MORE AVAILABLE IN ATP13A2 WT--------
ann_pct_wt_simple <- atac_clean %>%
  filter(M < 0) %>%
  count(annotation_simple, name = "n") %>%
  mutate(pct = 100 * n / sum(n)) %>%
  mutate(annotation_simple = factor(annotation_simple, levels = names(ann_colors_simple)))

ann_pct_wt_simple

ggplot(ann_pct_mut_simple, aes(x = pct, y = annotation_simple, fill = annotation_simple)) +
  geom_col(height = 0.7) +
  scale_fill_manual(values = ann_colors_simple) +
  labs(x = "% of peaks more accessible in ATP13A2 c.1306", y = NULL) +
  theme_classic() +
  theme(legend.position = "none")
#--------COUNTS DA PROMOTERS--------
promoter_pc_counts <- atac_promoters_best %>%
  mutate(direction = ifelse(M > 0, "More open in ATP13A2", "More open in WT")) %>%
  count(direction)

promoter_pc_counts
```

```{r setup, include=FALSE}
#---------INTEGRATION OF ATAC AND RNA SEQ -----------
deg_clean <- c.1306_DEG_TOP %>%
  rename(
    gene_symbol = gene,
    RNA_logFC   = logFC,
    RNA_PValue  = PValue
  ) %>%
  mutate(
    RNA_FDR = p.adjust(RNA_PValue, method = "BH")
  )

multiome <- atac_promoters_best %>%
  select(gene_symbol, ATAC_M = M, ATAC_A = A) %>%
  inner_join(deg_clean, by = "gene_symbol") %>%
  mutate(
    concordance = case_when(
      ATAC_M > 0 & RNA_logFC > 0 ~ "Concordant Up",
      ATAC_M < 0 & RNA_logFC < 0 ~ "Concordant Down",
      TRUE                       ~ "Discordant"
    )
  )

table(multiome$concordance)
N <- 15

top_up <- multiome %>%
  filter(concordance == "Concordant Up") %>%
  arrange(desc(ATAC_M + RNA_logFC)) %>%
  slice_head(n = N) %>%
  pull(gene_symbol)

top_down <- multiome %>%
  filter(concordance == "Concordant Down") %>%
  arrange(ATAC_M + RNA_logFC) %>%
  slice_head(n = N) %>%
  pull(gene_symbol)

top_concordant_genes <- unique(c(top_up, top_down))
top_concordant_genes

#------- FIGURE 4I----------- 
ggplot(multiome, aes(x = ATAC_M, y = RNA_logFC, color = concordance)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  ggrepel::geom_text_repel(
    data = subset(multiome, gene_symbol %in% top_concordant_genes),
    aes(label = gene_symbol),
    size = 3,
    color = "black"
  ) +
  scale_color_manual(values = c(
    "Concordant Up"   = "brown2",
    "Concordant Down" = "royalblue2",
    "Discordant"      = "grey60"
  )) +
  theme_bw() +
  labs(
    title = "ATAC vs RNA concordance (top concordant genes highlighted)",
    x = "ATAC promoter log2FC (ATP13A2 / WT)",
    y = "RNA log2FC (ATP13A2 / WT)"
  )

#----- HIGHLIGHT GENES-------
ggplot(multiome, aes(x = ATAC_M, y = RNA_logFC, color = concordance)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  ggrepel::geom_text_repel(
    data = subset(multiome, gene_symbol %in% astro_inflam_senescence_genes),
    aes(label = gene_symbol),
    size = 3,
    color = "black"
  ) +
  scale_color_manual(values = c(
    "Concordant Up"   = "brown2",
    "Concordant Down" = "royalblue2",
    "Discordant"      = "grey60"
  )) +
  theme_bw() +
  labs(
    title = "ATAC vs RNA concordance scatterplot",
    x = "ATAC promoter log2FC (ATP13A2 / WT)",
    y = "RNA log2FC (ATP13A2 / WT)"
  )

```